Fluent statistical computing interfaces for biological data analysis

### 2020-07-07 Rmd/01-intro.Rmd
rejig intro

This thesis focuses on core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation, with a focus 
on applications to genomics and transcriptomics. It is organised in 
the same manner that you would iterate back and
forth between these tasks during an analysis. To begin you wrangle your data 
into a structure that you can understand and use; section \@ref(sec:tidygranges)
introduces a grammar-based framework for transforming genomics data that is 
described in chapter \@ref(ch:plyranges). Next, you integrate data and model 
outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
introduces a framework for incorporating and discovering 'interesting' genomic
regions, described further in chapters \@ref(ch:fluentGenomics) and 
\@ref(ch:intron). Third, you cycle between model-based representations and 
visualisations of your data; section \@ref(sec:va) introduces a visual analytics 
approach for understanding non-linear dimension reduction techniques and is 
described in detail in chapter \@ref(ch:tsne). Lastly, you take a step back and 
look at the implications and limitations of your work; in chapter 
\@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
research.

@@

Chapter \@ref(ch:plyranges) shows that the *tidy data* semantic is applicable
to range-based genomics data and develops a *fluent* interface to transforming
it called *plyranges*. The software provides a framework to an assist
an analyst to compose queries on genomics datasets. This chapter has been 
published in Genome Biology [@Lee2019].

@@

It is rare that a biological data analysis will involve
a single measurement assay or that only one aspect of a measurement assay
will be of interest to the biological question under study. While there are 
many approaches to integrating data sets from multiple assays using multivariate
statistical techniques [@Meng2016-pz], in chapter \@ref(ch:fluentGenomics)
we describe a simpler end-to-end workflow for integrating 
results along the genome using *plyranges*. This workflow also shows that
our grammar based approach does not impair interoperability between
the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
together.

Moreover, in chapter \@ref(ch:intron) we explore the limits of the *tidy data*
semantic by extending *plyranges* to analyse coverage estimated on RNA-seq
data by developing a new software tool called *superintronic*. We show that the
long-form tidy representation is an effective way of combining the experimental
design and reference annotations into a single genomic data structure for
exploration. We use *superintronic* to develop a framework for discovering
interesting regions of coverage and apply our approach to integrating intron
signal from RNA-seq data.

@@

Chapter \@ref(ch:tsne) introduces
a software package called *liminal* for constructing these views and
a user interaction framework for identifying distortions. We present several
case studies for using our approach to diagnose the quality of 
results obtained via t-distributed stochastic neighbour embeddings (t-SNE)

>>>>

This thesis focuses on core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation, with a focus 
on applications to genomics and transcriptomics. It is organised in 
according to the aforementioned workflow steps. To begin we discuss wrangling
biological data using a coherent represntation and programming interface; 
section \@ref(sec:tidygranges)
introduces a grammar-based framework for transforming genomics data that is 
described in chapter \@ref(ch:plyranges). Next, we look at integrating data and 
model outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
introduces a framework for incorporating and discovering 'interesting' genomic
regions, described further in chapters \@ref(ch:fluentGenomics) and 
\@ref(ch:intron). Third, we consider the challenges in visualising  
high dimensional data; section \@ref(sec:va) introduces a visual analytics 
approach for understanding non-linear dimension reduction techniques described
in chapter \@ref(ch:tsne). Lastly, in chapter 
\@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
developments.

@@

Chapter \@ref(ch:plyranges) shows that the *tidy data* semantic is applicable
to range-based genomics data and develops a *fluent* interface to transforming
it called *plyranges*. The software provides a framework to an assist
an analyst to compose queries on genomics datasets. This chapter has been 
published in @Lee2019.

@@

It is rare that a biological data analysis will involve
a single measurement assay or that only one aspect of a measurement assay
will be of interest to the biological question under study. While there are 
many approaches to integrating data sets from multiple assays using multivariate
statistical techniques [@Meng2016-pz], in chapter \@ref(ch:fluentGenomics)
we describe a simple end-to-end workflow for integrating 
results along the genome using *plyranges*. This workflow also shows that
our grammar based approach does not impair interoperability between
the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
together. This chapter has been published in @Lee2020-ie.

Moreover, in chapter \@ref(ch:intron) we explore the limits of the *tidy data*
semantic by extending *plyranges* to analyse coverage estimated on RNA-seq
data by developing a new software tool called *superintronic*. We show that the
long-form tidy representation is an effective way of combining the experimental
design and reference annotations into a single genomic data structure for
exploration. We use *superintronic* to develop a framework for discovering
interesting regions of coverage and apply our approach to integrating intron
signal from RNA-seq data. This chapter is based on the preprint @Lee2019-mf.

@@

Chapter \@ref(ch:tsne) introduces
a software package called *liminal* for constructing these views and
a user interaction framework for identifying distortions. We present several
case studies using data sets from both single-cell transciptomics and 
machine learning  for using our approach to diagnose the quality of 
results obtained via popular non-linear DR methods
like t-distributed stochastic neighbour embeddings (t-SNE).
